---
title: "IUNC Red List Assessment: Polar Bear"
subtitle: "Report"
author: "Ruowen Xiao"
date: 2025-10-04
format: html
toc: true  # enable Table of Contents
code-fold: true
message: false
warning: false
execute: 
  echo: false  # do not show r chunks in the generated pdf
---

## 1. Introductions

## 2. Methods

t, ice, N_2015, N_t, B_i

```{r}
library(ggplot2)
```

### 2.1. Generation Length

> Modification: They propagate percentiles, but you are to use a probabilistic approach to take into account uncertainties in the assessment.

`GL`: Generation Length (years)

Table 1. Mean `GL` estimates for 11 subpopulations(mean, lower 95% CI, upper 95% CI)

The mean of subpopulation mean estimates was 11.5 years (95% CI = 9.8, 13.6).

```{r}
gl_mean = 11.5
gl_lower = 9.8
gl_upper = 13.6
# gl_N = 3374

gl_sd = (gl_upper - gl_lower) / qnorm(0.975) / 2
```

```{r}
# Prepartion for the MC
gl_samples <- function(niter) {
  return(rnorm(niter, gl_mean, gl_sd))
}

hist(gl_samples(1000))
```

### 2.2. Sea Ice

> Modification: You are to consider recent projections on the Last Ice Area.
>
> [Last Ice Area](https://www.arcticwwf.org/our-priorities/last-ice-area/)

`ice`: reduced-ice days(days)

`T`: the threshold area()

The threshold area (denoted T) was chosen as the midpoint of these values:

### 2.3. Population Projections

#### 2.3.1. Abundance Estimates

Table 3. Abundance(`N`) estimates for 19 subpopulations

#### 2.3.2. Statistical Models and Computer Simulation

Projections started in the year 2015 and ended in year t = 2015 + (3 × GL).

Normalization of abundance was performed separately for each subpopulation `i`:

$$
N^{norm}_i = \frac{N_i}{N^1_i}
$$

$N_i$ is either the first or second available abundance estimate

$N^1_i$ is the first available abundance estimate.

The linear model for Approach 2 included an intercept for each subpopulation and a single, global slope coefficient：

$$
N^{norm} = \Sigma_i B_i + B_{Global} \cdot fitted.ice + \epsilon
$$

For each draw of starting abundance for subpopulation i, we predicted mean abundance in year t as:

$$
N_{i,t} = N_{i,2015} + \Delta N_{i,t} \times N_{i,2015}
$$

Finally, we calculated percent change in mean global population size as:

$$
\Delta G = 100 \times \frac{G_t - G_2015}{G2015}
$$

```{r}
niter <- 62500

df_par <- data.frame(gl = gl_samples(niter),
                     iter = 1:niter)
```

## 3. Results

## 4. Conclusions

> Modification: You are also to assess uncertainty in the conclusion which could be different from the conclusion in the Red List Assessment.

Overall uncertainty

## 5. Communication

Discuss the importance of taking action to stop the decline in the Last Ice Area to protect polar bear.

Target audience is policy makers

Communicate the conclusion, overall uncertainty and the path leading to the conclusion in your text.
